---
title: "Root mean squared error"
output: html_document
---

## Monte Carlo simulation study 

### Overview: 

1. Simulate cell type-specific DNAm profiles with 
platform-dependent random effects ($X_k$ matrices)
with different mean methylation levels 
(alpha0, alpha1, sig0, sig1) varying across two platforms. 
These matricies are used for all simulated samples in Step 2. 
2. For a given DNAm sample (1 to `nSamples` e.g. 100) 
  - Simulate a true cell composition $\pi$. 
  - Simulate a 450k array sample (Y450k = X450k * pi + err) 
     and RRBS sample (Ybs = Xbs * pi + err)
3. Estimate cell composition 450k and RRBS samples using 
Houseman and methylCC methods. Scale cell composition
estimates to sum to 1, if needed.
4. Calculate $RMSE_k$ for each cell type $k$. 
Average across cell types for mean RMSE. 

### Run code on JHPCE

```{r}
library(minfi)
library(quadprog)
library(methylCC)
library(doParallel)
library(abind)
library(tidyr)
library(truncnorm)

dataPath <- "/users/shicks1/projects/methylCC/simStudy"

# Z matrix with 64 regions and 4 cell types: 0 = unmethylated, 1 = methylated
Z = as.matrix(expand.grid(c(0,1), c(0,1), c(0,1), c(0,1), c(0,1), c(0,1)))
Z = Z[-c(1, 64),]
Z = rbind(# rep(0, 6), 
          Z[c(1,2,4,8,16,32,31,47,55,59,61,62), ])#, 
          # rep(1, 6))

Z = apply(Z, 2, function(x){rep(x, each=4)})
IDs = c("CD8T","CD4T", "NK","Bcell","Mono","Gran")
colnames(Z) <- IDs
K = ncol(Z)
R = nrow(Z)
nSamples = 40 # number of whole blood samples
sdtrue = 0.05

alpha0450k = 0.15; alpha1450k = 0.75
sd0450k = 0.03; sd1450k = 0.04
alpha0bs = 0.03; alpha1bs = 0.85
sd0bs = 0.01; sd1bs = 0.03

N <- 20

nCores = 1
registerDoParallel(cores = nCores)
workers <- getDoParWorkers()
backend <- getDoParName()
version <- getDoParVersion()

Sys.time()
simResults <- foreach(kk = 1:N) %dopar% {
  
  delta0450k = matrix(rnorm(R, mean = alpha0450k, sd = sd0450k), ncol = 1)
  delta1450k = matrix(rnorm(R, mean = alpha1450k, sd = sd1450k), ncol = 1)
  X450k = (sweep((1-Z), 1, delta0450k, FUN = "*")) + (sweep(Z, 1, delta1450k, FUN = "*"))
  X450k[X450k<0] <- 0.01; X450k[X450k>1] <- 0.99
  
  delta0bs = matrix(rnorm(R, mean = alpha0bs, sd = sd0bs), ncol = 1)
  delta1bs = matrix(rnorm(R, mean = alpha1bs, sd = sd1bs), ncol = 1)
  Xbs = (sweep((1-Z), 1, delta0bs, FUN = "*")) + (sweep(Z, 1, delta1bs, FUN = "*"))
  Xbs[Xbs<0] <- 0.01; Xbs[Xbs>1] <- 0.99
  
  simResultsSub <- array(dim=c(K, 5, nSamples))
  pi.truth <- array(0, dim=c(K, nSamples))
  pi.truth[1,] <- rtruncnorm(nSamples, a = 0, mean = 0.036, sd = 0.040) # CD8T
  pi.truth[2,] <- rtruncnorm(nSamples, a = 0, mean = 0.180, sd = 0.078) # CD4T
  pi.truth[3,] <- rtruncnorm(nSamples, a = 0, mean = 0.076, sd = 0.052) # NK
  pi.truth[4,] <- rtruncnorm(nSamples, a = 0, mean = 0.040, sd = 0.049) # Bcell
  pi.truth[5,] <- rtruncnorm(nSamples, a = 0, mean = 0.095, sd = 0.031)   # Mono 
  pi.truth[6,] <- rtruncnorm(nSamples, a = 0, mean = 0.574, sd = 0.109) # Gran
  pi.truth <- sweep(abs(pi.truth), 2, colSums(abs(pi.truth)), FUN = "/")
    
  err = matrix(rnorm(R*nSamples, mean = 0, sd = sdtrue), nrow=R, ncol=nSamples)
  Y450k <- abs(X450k %*% pi.truth + err)
    
  err = matrix(rnorm(R*nSamples, mean = 0, sd = sdtrue), nrow=R, ncol=nSamples)
  Ybs = abs(Xbs %*% pi.truth + err)
    
  #### 450k data
  # Houseman method
  mleHouse450k <- minfi:::projectCellType(Y450k, X450k)
  mleHouse450k <- sweep(mleHouse450k, 1, rowSums(mleHouse450k), FUN = "/")
    
  # Hicks method
  est <- estimateCC(object=Y450k, regionMat=Z, initParamMethod = "random",
                    epsilon = .1, verbose = FALSE, maxIter = 50)
  mleHicks450k <- as.matrix(cellcounts(est))
    
  # Houseman method
  mleHousebs = minfi:::projectCellType(Ybs, X450k)
  mleHousebs <- sweep(mleHousebs, 1, rowSums(mleHousebs), FUN = "/")
    
  # Hicks method
  est <- estimateCC(object=Ybs, regionMat=Z, initParamMethod = "random",
                    epsilon = .1, verbose = FALSE, maxIter = 50)
  mleHicksbs <- as.matrix(cellcounts(est))
    
  simResultsSub[,1,] <- pi.truth
  simResultsSub[,2,] <- t(mleHouse450k)
  simResultsSub[,3,] <- t(mleHicks450k)
  simResultsSub[,4,] <- t(mleHousebs)
  simResultsSub[,5,] <- t(mleHicksbs)
  
  return(simResultsSub)
}
Sys.time()

simResults <- abind(simResults, along = 4)

message("Saving results.")
saveRDS(simResults, file=file.path(dataPath, "simData/simResults.rds"))
```

### Create Figure S1

```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(cowplot)
library(tidyr)

dataPath <- "/users/shicks1/projects/methylCC/simStudy"
Z = as.matrix(expand.grid(c(0,1), c(0,1), c(0,1), c(0,1), c(0,1), c(0,1)))
Z = Z[-c(1, 64),]
Z = rbind(# rep(0, 6), 
          Z[c(1,2,4,8,16,32,31,47,55,59,61,62), ])#, 
          # rep(1, 6))

Z = apply(Z, 2, function(x){rep(x, each=4)})
colnames(Z) <- IDs
K = ncol(Z)
R = nrow(Z)
sdtrue = 0.1

alpha0450k = 0.15; alpha1450k = 0.75
sd0450k = 0.03; sd1450k = 0.04
alpha0bs = 0.03; alpha1bs = 0.85
sd0bs = 0.01; sd1bs = 0.03

# simResults <- readRDS(file.path(dataPath, "simData/simResults.rds")) 
nSamples <- dim(simResults)[3]  # number of whole blood samples
N <- dim(simResults)[4] # number of MC simulations

# calculate RMSE
house450k.all = hicks450k.all = housebs.all = hicksbs.all = array(dim=c(N, length(c("bias","mse"))))
for(kk in 1:N){
  
  house450k = data.frame("rmse" = sqrt(t(t(rowMeans((simResults[,2,,kk] - simResults[,1,,kk])^2)))))
  hicks450k = data.frame("rmse" = sqrt(t(t(rowMeans((simResults[,3,,kk] - simResults[,1,,kk])^2)))))
  housebs = data.frame("rmse" = sqrt(t(t(rowMeans((simResults[,4,,kk] - simResults[,1,,kk])^2)))))
  hicksbs = data.frame("rmse" = sqrt(t(t(rowMeans((simResults[,5,,kk] - simResults[,1,,kk])^2)))))

  # average across cell types
  house450k.all[kk,] <- colMeans(house450k)
  hicks450k.all[kk,] <- colMeans(hicks450k)
  housebs.all[kk,] <- colMeans(housebs)
  hicksbs.all[kk,] <- colMeans(hicksbs)
}

# pdf(file.path(dataPath, "figs/FigureS1.pdf"), width = 12, height = 4)
par(mfrow=c(1,3))
xs = seq(0,1,by=0.001)
plot(x=xs, y = dnorm(x=xs, mean = alpha0450k, sd = sd0450k), type = "l", 
     ylim = c(0, 45), lwd = 2, cex.axis = 1.5, cex.lab = 1.5, cex.main = 1.5,
     xlab = "Methylation", ylab = "density", 
     main = "Simulated platform-dependent\nrandom effects")
lines(x=xs, y = dnorm(x=xs, mean = alpha0bs, sd = sd0bs), type = "l", 
      lty = 2, lwd = 2)
lines(x=xs, y = dnorm(x=xs, mean = alpha1450k, sd = sd1450k), 
      type = "l", lwd = 2)
lines(x=xs, y = dnorm(x=xs, mean = alpha1bs, sd = sd1bs), type = "l", 
      lty = 2, lwd = 2)
legend('topright', c("450k", "RRBS"), col = 1, lty = 1:2, lwd = 3, cex = 1.5)
mtext(text = "A", side = 3, line = 0, cex = 2, adj = 0, at = -0.15)

boxplot(cbind("Houseman" = house450k.all[,1], "methylCC" = hicks450k.all[,1]), 
        main = "Simulated data from 450k platform", ylim=c(0.0, 0.09), 
        xlab="Cell composition estimation method", 
        ylab = "RMSE", cex.axis = 1.5, cex.lab = 1.5, cex.main = 1.5)
mtext(text = "B", side = 3, line = 0, cex = 2, adj = 0, at = 0.05)
boxplot(cbind("Houseman" = housebs.all[,1], "methylCC" = hicksbs.all[,1]), 
        main = "Simulated data from RRBS platform", ylim=c(0.00, 0.09), 
        xlab="Cell composition estimation method", 
        ylab = "RMSE", cex.axis = 1.5, cex.lab = 1.5, cex.main = 1.5)
mtext(text = "C", side = 3, line = 0, cex = 2, adj = 0, at = 0.05)
# dev.off()

```

