---
title: Summary of BLUEPRINT Analyses
author: Stephanie Hicks
date: "August 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(dplyr)
library(ggplot2)
library(cowplot)
library(bsseq)

workingDir_blueprint <- 
  "/users/shicks1/projects/methylCCPaper/case-studies/blueprint-wgbs"
dataPath <- "/users/shicks1/data/DNAm/blueprint_ihec"
```

```{r}
custom_table <- readRDS(file = file.path(dataPath,"blueprint_blood_custom_table.RDS"))
custom_table$file_name <- unlist(lapply(stringr::str_split(custom_table$file_path, "CNAG/"), 
                                        function(x){ x[[2]]}))
custom_table$biosource_short <- unlist(lapply(stringr::str_split(custom_table$biosource, "-"), 
                                              function(x){ x[[1]]}))
```

## Identified cell type specific CpGs; purified cell types from 450k platform 

Cell composition estimates from purified whole blood celltypes
measured on WGBS platform, but cell type-specific CpGs identified 
using purified cell types measured on 450K platform

```{r}
counts_methylcc <- readRDS(file.path(dataPath, "files_cellcomp",
                              "blueprint_blood_cc_methylcc-only-cpgs.RDS"))

custom_table_sub <- custom_table %>% 
  filter(grepl("calls.bs_call", name))

custom_table_ordered <- custom_table_sub[match(rownames(counts_methylcc), 
                                               custom_table_sub$sample_name), ]  
custom_table_ordered$cell_type <-
  factor(plyr::revalue(custom_table_ordered$biosource_short,
                       c("CD14"="Mono", "CD38"="Bcell", "CD4" = "CD4T",
                         "CD8"="CD8T", "cytotoxic CD56" = "NK",
                         "mature eosinophil" = "Gran",
                         "mature neutrophil" = "Gran")))
xmat = model.matrix(~ custom_table_ordered$cell_type - 1)
colnames(xmat) = levels(custom_table_ordered$cell_type)

df_methylcc <- 
  tidyr::gather(cbind("samples" = rownames(counts_methylcc), 
                      "biosource" = custom_table_ordered$biosource,
                      "biosource_short" = custom_table_ordered$biosource_short,
                      counts_methylcc), 
                celltype, est, -samples, -biosource, -biosource_short)

df_methylcc$biosource <- 
  factor(df_methylcc$biosource,
         levels = c("CD4-positive, alpha-beta T cell",
                    "CD8-positive, alpha-beta T cell", 
                    "mature neutrophil", "mature eosinophil", 
                    "CD38-negative naive B cell", 
                    "CD14-positive, CD16-negative classical monocyte", 
                    "cytotoxic CD56-dim natural killer cell"))

new_levels <- c("CD4-positive,\nalpha-beta T cell (N=12)", 
                "CD8-positive,\nalpha-beta T cell (N=4)",
                "mature neutrophil (N=12)", "mature eosinophil (N=2)",
                "CD38-negative\nnaive B cell (N=4)", 
                "CD14-positive, CD16-negative\nclassical monocyte (N=6)", 
                "cytotoxic CD56-dim\nnatural killer cell (N=4)")
levels(df_methylcc$biosource) <- new_levels

gmat_cpg <- df_methylcc %>% 
  ggplot(aes(x = biosource, y = est)) + 
  geom_jitter(size=3, width = 0.25, aes(color = celltype)) + 
  coord_flip() + 
  xlab("cell types") +
  ylab("methylCC") +
  labs(title = "Cell composition estimates from purified celltypes measured\non WGBS platform (N=44, data from BLUEPRINT)\n(cell type-specific CpGs identified using\npurified cell types measured on 450K platform)", 
         color = "Cell type") + 
  theme(legend.position = "top", legend.justification= "center") + 
  scale_color_discrete(name = "Cell type", 
                       labels = c("B-cells", "CD4 T-cells", "CD8 T-cells", 
                                  "Granulocytes","Monocytes", "Natural killer cells")) 

pdf(file.path(workingDir_blueprint, "figs/fig-blueprint-ests-only-cpgs.pdf"), 
    width = 9, height = 7)
print(gmat_cpg)
dev.off()

```

## Identified cell type specific DMRs; purified cell types from 450k platform 

Cell composition estimates from purified whole blood celltypes
measured on WGBS platform, but cell type-specific DMRs identified 
using purified cell types measured on 450K platform

```{r}
counts_methylcc <- readRDS(file.path(dataPath, "files_cellcomp",
                              "blueprint_blood_cc_methylcc.RDS"))
custom_table_sub <- custom_table %>% 
  filter(grepl("calls.bs_call", name))

custom_table_ordered <- custom_table_sub[match(rownames(counts_methylcc), 
                                               custom_table_sub$sample_name), ]  
custom_table_ordered$cell_type <-
  factor(plyr::revalue(custom_table_ordered$biosource_short,
                       c("CD14"="Mono", "CD38"="Bcell", "CD4" = "CD4T",
                         "CD8"="CD8T", "cytotoxic CD56" = "NK",
                         "mature eosinophil" = "Gran",
                         "mature neutrophil" = "Gran")))
xmat = model.matrix(~ custom_table_ordered$cell_type - 1)
colnames(xmat) = levels(custom_table_ordered$cell_type)

df_methylcc <- 
  tidyr::gather(cbind("samples" = rownames(counts_methylcc), 
                      "biosource" = custom_table_ordered$biosource,
                      "biosource_short" = custom_table_ordered$biosource_short,
                      counts_methylcc), 
                celltype, est, -samples, -biosource, -biosource_short)

df_methylcc$biosource <- 
  factor(df_methylcc$biosource,
         levels = c("CD4-positive, alpha-beta T cell",
                    "CD8-positive, alpha-beta T cell", 
                    "mature neutrophil", "mature eosinophil", 
                    "CD38-negative naive B cell", 
                    "CD14-positive, CD16-negative classical monocyte", 
                    "cytotoxic CD56-dim natural killer cell"))

new_levels <- c("CD4-positive,\nalpha-beta T cell (N=12)", 
                "CD8-positive,\nalpha-beta T cell (N=4)",
                "mature neutrophil (N=12)", "mature eosinophil (N=2)",
                "CD38-negative\nnaive B cell (N=4)", 
                "CD14-positive, CD16-negative\nclassical monocyte (N=6)", 
                "cytotoxic CD56-dim\nnatural killer cell (N=4)")
levels(df_methylcc$biosource) <- new_levels

gmat_dmr <- df_methylcc %>%
  ggplot(aes(x = biosource, y = est)) + 
  geom_jitter(size=3, width = 0.25, aes(color = celltype)) + 
  coord_flip() + 
  xlab("cell types") +
  ylab("methylCC") +
  labs(title = "Cell composition estimates from purified celltypes measured\non WGBS platform (N=44, data from BLUEPRINT)\n(cell type-specific DMRs identified using\npurified cell types measured on 450K platform)", 
         color = "Cell type") + 
  theme(legend.position = "top", legend.justification= "center") + 
  scale_color_discrete(name = "Cell type", 
                       labels = c("B-cells", "CD4 T-cells", "CD8 T-cells", 
                                  "Granulocytes","Monocytes", "Natural killer cells"))  + 
  geom_text(x=6, y=0.5, size = 5, fontface = "bold",
            label=paste0("RMSE: ", round(mean(sqrt(colMeans((counts_methylcc - xmat)^2))),3)))


pdf(file.path(workingDir_blueprint, "figs/fig-blueprint-ests-new.pdf"), 
    width = 9, height = 7)
print(gmat_dmr)
dev.off()

```

## Identified cell type specific DMRs; purified cell types from WGBS platform 

Cell composition estimates from purified whole blood celltypes
measured on WGBS platform, but cell type-specific DMRs identified 
using purified cell types measured on WGBS platform

```{r}
counts_methylcc <- readRDS(file.path(dataPath, "files_cellcomp",
                              "blueprint_blood_cc_methylcc_wgbsregions.RDS"))

custom_table_sub <- custom_table %>% 
  filter(grepl("calls.bs_call", name))

custom_table_ordered <- custom_table_sub[match(rownames(counts_methylcc), 
                                               custom_table_sub$sample_name), ]  

custom_table_ordered$cell_type <-
  factor(plyr::revalue(custom_table_ordered$biosource_short,
                       c("CD14"="Mono", "CD38"="Bcell", "CD4" = "CD4T",
                         "CD8"="CD8T", "cytotoxic CD56" = "NK",
                         "mature eosinophil" = "Gran",
                         "mature neutrophil" = "Gran")))
xmat = model.matrix(~ custom_table_ordered$cell_type - 1)
colnames(xmat) = levels(custom_table_ordered$cell_type)

df_methylcc <- 
  tidyr::gather(cbind("samples" = rownames(counts_methylcc), 
                      "biosource" = custom_table_ordered$biosource,
                      "biosource_short" = custom_table_ordered$biosource_short,
                      counts_methylcc), 
                celltype, est, -samples, -biosource, -biosource_short)

df_methylcc$biosource <- 
  factor(df_methylcc$biosource,
         levels = c("CD4-positive, alpha-beta T cell",
                    "CD8-positive, alpha-beta T cell", 
                    "mature neutrophil", "mature eosinophil", 
                    "CD38-negative naive B cell", 
                    "CD14-positive, CD16-negative classical monocyte", 
                    "cytotoxic CD56-dim natural killer cell"))

new_levels <- c("CD4-positive,\nalpha-beta T cell (N=12)", 
                "CD8-positive,\nalpha-beta T cell (N=4)",
                "mature neutrophil (N=12)", "mature eosinophil (N=2)",
                "CD38-negative\nnaive B cell (N=4)", 
                "CD14-positive, CD16-negative\nclassical monocyte (N=6)", 
                "cytotoxic CD56-dim\nnatural killer cell (N=4)")
levels(df_methylcc$biosource) <- new_levels

gmat_dmr_wgbs <- df_methylcc %>% 
  ggplot(aes(x = biosource, y = est)) + 
  geom_jitter(size=3, width = 0.25, aes(color = celltype)) + 
  coord_flip() + 
  xlab("cell types") +
  ylab("methylCC") +
  labs(title = "Cell composition estimates from purified celltypes measured\non WGBS platform (N=44, data from BLUEPRINT)\n(cell type-specific DMRs identified using\npurified cell types measured on WGBS platform)", 
         color = "Cell type") + 
  theme(legend.position = "top", legend.justification= "center") +
  scale_color_discrete(name = "Cell type", 
                       labels = c("B-cells", "CD4 T-cells", "CD8 T-cells", 
                                  "Granulocytes","Monocytes", "Natural killer cells")) + 
  geom_text(x=6, y=0.5, size = 5, fontface = "bold",
            label=paste0("RMSE: ", round(mean(sqrt(colMeans((counts_methylcc - xmat)^2))),3)))


pdf(file.path(workingDir_blueprint, "figs/fig-blueprint-ests-wgbs-regions-new.pdf"), 
    width = 9, height = 7)
print(gmat_dmr_wgbs)
dev.off()

```

## Combine figures
  
```{r}
p1 <- plot_grid(gmat_dmr, gmat_dmr_wgbs, ncol = 1, 
                labels = LETTERS[1:2], label_size = 30)

pdf(file.path(workingDir_blueprint, "figs/fig-blueprint-ests-three.pdf"), 
    width = 10, height = 12)
print(p1)
dev.off()

```



## Exploring cell type-specific DMRs in 450K vs WGBS purified whole blood samples 

```{r}
data_tidy <- readRDS(file.path(dataPath, "blueprint_celltype-450k-wgbs-withCpGs.RDS"))

pdf("/users/shicks1/projects/methylCCPaper/case-studies/blueprint-wgbs/figs/test.pdf", width=6, height = 6)
data_tidy %>% 
  ggplot(aes(x = X1, y = X2, color = .id)) + 
  geom_point() + 
  xlab("450K platform") +
  ylab("WGBS platform") +
  labs(title = "Methylation levels in cell type-specific DMRs\nusing purified whole blood cell types", 
       color = "Cell type") +
  theme(legend.position = "top", legend.justification= "center") + 
  scale_color_discrete(name = "Cell type", 
                       labels = c("B-cells", "CD4 T-cells", "CD8 T-cells", 
                                  "Granulocytes","Monocytes", "Natural killer cells"))
dev.off()

```



```{r}
data_tidy <- readRDS(file.path(dataPath, "blueprint_celltype-450k-wgbs-noCpGs.RDS"))

pdf("/users/shicks1/projects/methylCCPaper/case-studies/blueprint-wgbs/figs/fig-comparison-profiles-450k-wgbs.pdf", width=6, height = 6)
data_tidy %>% 
  ggplot(aes(x = X1, y = X2, color = .id)) + 
  geom_point() + 
  xlab("450K platform") +
  ylab("WGBS platform") +
  labs(title = "Methylation levels of purified whole blood cell types\n(cell type-specific DMRs identified\nusing purified cell types measured on 450K platform)", 
       color = "Cell type") +
  theme(legend.position = "top", legend.justification= "center") + 
  scale_color_discrete(name = "Cell type", 
                       labels = c("B-cells", "CD4 T-cells", "CD8 T-cells", 
                                  "Granulocytes","Monocytes", "Natural killer cells")) + 
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "gray") + 
  geom_vline(xintercept = 0.5, linetype = "dashed", colour = "gray")
dev.off()

```

## Exploring WGBS DMRs

```{r}
library(rhdf5) 
library(HDF5Array)
library(bsseq)
library(dmrseq)

hdf5_bs_se_path <- file.path(dataPath, "files_bsseq_hdf5_se")
bs <- loadHDF5SummarizedExperiment(hdf5_bs_se_path)

IDs = c("Gran", "CD4T", "CD8T", "Bcell","Mono", "NK")

# extract beta values, phenotypic information and GRanges objects
cell <- factor(bs$cell_type, levels = IDs)
cell_levels <- levels(cell)

# define design matrix to search for DMRs
xmat = model.matrix(~ cell - 1)
colnames(xmat) = cell_levels
pData(bs) <- cbind(pData(bs), xmat)

dmrs_found <- find_dmrs_wgbs(num_regions = 50, dmr_up_cutoff = 1, 
                              dmr_down_cutoff = 1)

pdf(file.path(workingDir_blueprint, 
              "figs/fig-blueprint-coefs.pdf"), width = 10, height = 6)
par(mfrow=c(1,2))
image(t(dmrs_found$y_regions), xaxt="n", ylab = "regions", 
      main = "WGBS cell type-specific DMRs\n(N=44 samples)")
axis(side=1, at=seq(0, 1, length.out=44), labels=FALSE)
text(seq(0, 1, length.out=44), par("usr")[3] - 0.1, 
     labels = dmrs_found$pd$cell_type, srt = 90, pos = 1, xpd = TRUE)

image(t(dmrs_found$profiles), xaxt="n", ylab = "regions", 
      main = "WGBS cell type-specific profiles\n(avg across cell types)")
axis(side=1, at=seq(0, 1, by =.2), labels=colnames(dmrs_found$zmat))
dev.off()

regs <- vector("list", length(IDs))
for(ind in seq_len(length(IDs))){
  regs[[ind]] <- readRDS(file = file.path(dataPath, "files_dmrs", 
                              paste0("blueprint_blood_regions_dmrseq_", 
                                     tolower(IDs[ind]),".RDS")))
}

# get annotations for hg18
library(annotatr)
annoTrack <- dmrseq::getAnnot("hg19")

pdf(file.path(workingDir_blueprint, 
              "figs/fig-blueprint-top-wgbs-regions.pdf"), width = 10, height = 6)
for(ind in seq_len(length(IDs))){
dmrseq::plotDMRs(bs, regions=regs[[ind]][1:5,], testCovariate=IDs[ind],
         annoTrack=annoTrack, main = paste0("top 5 ", IDs[ind], " cell type-specific DMRs"))
}
dev.off()


```

